---
title: "Exploring the data.table package"
author: "Reto Schmucki"
date: "29 March 2020"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
---

### Using data.table R package

Under the hood, many function of operation included in the rbms package are done using the `data.table` package. The <a href="https://cloud.r-project.org/web/packages/data.table/vignettes/datatable-intro.html">data.table</a> package offer fast and powerful operation on large data set with a very compact syntax. 

---

In this session, you will:

1. read csv file with data.table
2. merge data set with one or multiple key
3. rename column names
4. extract basic statistics

----

You will need to load the following R packages and the data found in the bms_workshop_data folder that can be <a href="https://filedn.com/l4iF16owVCHBm81sGhdrcpX/bms_workshop_data.zip">downloaded here</a>.
Once you have downloaded the data, unzip the folder and add the data in your R project directory, your current R working directory, or set your working directly accordingly `setwd()`.

``` {r load_library}
library(data.table)
```

All data are stored in `rds` format, this format is highly efficient for storing R object, but you could also have them in any other format. To load `.rds data`, we use the function `readRDS()`

``` {r load_bms_data}
b_count_sub <- readRDS("my_code/bms_workshop_data/work_count.rds")
m_visit_sub <- readRDS("my_code/bms_workshop_data/work_visit.rds")
m_site_sub <- readRDS("my_code/bms_workshop_data/work_site.rds")

## to read csv into a data.table use fread()

bcz_class <- data.table::fread("my_code/bms_workshop_data/GEnS_v3_classification.csv")

b_count_sub
```

When you look at a data.table object, you can will see the first and last five row, with all column names. 

1. You can access specific column the same way as you would do with a data.frame with the \$ sign or a vector of names such as c("col1", "col2, ..., "col7"). 
2. You can subset the data.table with an logical argument of indices in the first position after the square bracket (e.g my_data_table[ col1 <= 453, ])
3. You can create new column with the ":=" symbol and a function (e.g. my_data_table[col1 <= 453, new_col := mean(col1)])
4. You can compute operation by group using a "by" argument in third position within the square brackets (e.g.  my_data_table[col1 <= 453, new_col := mean(col1), by = col3])

### Extract columns
``` {r manipulate_data.table1}
my_new_dt <- b_count_sub[ , .(bms_id, transect_id_serial, year, visit_date)]
my_new_dt
```
### Extract unique
``` {r manipulate_data.table_operation}
unique(my_new_dt$bms_id)
unique(my_new_dt[, .(bms_id, year)])
```

### Subset
``` {r manipulate_data.table_subset}
my_new_dt[bms_id == "NLBMS", ]
```
### Work with date
``` {r manipulate_data.table_date}
my_new_dt[ , month := month(visit_date)][ , c("day", "year") := .(mday(visit_date), year(visit_date))]
my_new_dt
```

### Count object
``` {r manipulate_data.table_countN}
my_new_dt[ , .N, by = bms_id]
```

### rename column
``` {r manipulate_data.table_rename}
## change names
setnames(b_count_sub, "transect_id_serial", "SITE_ID")
setnames(b_count_sub, c("species_name", "bms_id"), c("SPECIES", "BMS_ID"))
names(b_count_sub) <- toupper(names(b_count_sub))
b_count_sub
```

### Merge data sets
``` {r manipulate_data.table_merge}
## merge data set
setnames(m_site_sub, "transect_id_serial", "SITE_ID")

setkey(b_count_sub, SITE_ID)
setkey(m_site_sub, SITE_ID)

merge(b_count_sub, m_site_sub[bms_id == "FRBMS" , .(SITE_ID, transect_lon_1km, transect_lat_1km)])
merge(b_count_sub, m_site_sub[bms_id == "FRBMS" , .(SITE_ID, transect_lon_1km, transect_lat_1km)], all.x = TRUE)

data_m <- merge(b_count_sub, m_site_sub[ , .(SITE_ID, transect_lon_1km, transect_lat_1km)], all.x = TRUE)
data_m
```

There is many way to manipulate tabular data sets in R, dplyr and data.table are especially efficient with large data set, data.table can manipulate millions of row with ease. 
Learn one of these tool will help you seriously in every R task and analysis as 80% of the work is to organize the data to make them fit for the analysis or build a graph.
