---
title: "Mapping and extracting GIS information"
author: "Reto Schmucki"
update: "31 March 2025"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
---

### Mapping BMS sites

Working with spatially explicit data means that you can link them to various other environmental data sets. This also allow you to explore your data points visually and get a glimpse of there distribution and geographical context. In this short tutorial, we will explore how to generate spatial object from your BMS site data, using the [R package sf](https://r-spatial.github.io/sf/index.html), a powerful tool for geospatial computation in R. 

---

In this session, you will:

1. read csv file with data.table
2. convert coordinates in spatial object using sf
3. read raster file with the raster package
4. produce a map with rasterVis and tmap
5. extract value from raster using a vector object

----

You will need to load the following R packages and the data found in the bms_workshop_data folder that can be <a href="https://filedn.com/l4iF16owVCHBm81sGhdrcpX/bms_workshop_data.zip">downloaded here</a>.
Once you have downloaded the data, unzip the folder and add the data in your R project directory, your current R working directory, or set your working directly accordingly `setwd()`.

``` {r load_library}
library(data.table)
library(sf)
library(terra)
library(tmap)
library(mapview)
```

All data are stored in `rds` format, this format is highly efficient for storing R object, but you could also have them in any other format. To load `.rds data`, we use the function `readRDS()`

``` {r load_bms_data}
b_count_sub <- readRDS("bms_workshop_data/work_count.rds")
m_visit_sub <- readRDS("bms_workshop_data/work_visit.rds")
m_site_sub <- readRDS("bms_workshop_data/work_site.rds")
```
For this session, we will also upload a raster file with Bioclimate Regions provided by <a href="https://onlinelibrary.wiley.com/doi/abs/10.1111/geb.12022">Metzger et al. 2013</a> (see [licence](data/readme_metzger_data.txt)). tje raster file is provided in `.grd` format and can be read into R with the `raster()` function from the raster package. Together with the raster file with the Bioclimate Regions, you can load the classification available in a `.csv` and can be loaded with `data.table`. To subset the data from the eBMS database, I used a bounding box that can be loaded in R with the package `sf`.


### Mapping

``` {r load_spatial_data, results='hide', message=FALSE, warning=FALSE}
bcz <- terra::rast("bms_workshop_data/metzger_bcz_genz3.grd")
bb <- sf::st_read("bms_workshop_data/bbox.shp")
bcz_class <- data.table::fread("bms_workshop_data/GEnS_v3_classification.csv")
```

First we will produce a spatial object for the points corresponding to the BMS transects in the m_site_sub object (where "m" stands for monitoring). Using the sf function `st_as_sf()` you can build points from longitude/latitude coordinates and CRS=3035 as we use the standard **ETRS89-extended/LAEA Europe** as projection system [EPGS:3035](https://epsg.io/3035). You can have a quick look at the points on a map with the function `mapview()` from the R package mapview. This will open a window in your browser with an interactive map of your points. 

``` {r build_sf_object}
## sf transect location
site_sf <- sf::st_as_sf(m_site_sub, coords = c("transect_lon_1km", "transect_lat_1km"), crs = 3035)
mapview::mapview(site_sf)
```

> Trick: you can extract a bounding box with the function draw() from terra package and build a corresponding polygon with the function st_as_sfc() from sf package.
> `st_as_sfc(st_bbox(terra::draw(), crs = st_crs(4326)))`

Another option is to use the package tmap to produce "better looking map", but also much slower. For raster you will the function `tm_raster()` and add other layers using the same principle as in ggplot. Obviously, producing good looking and informative maps is an art and you can spend ages working on them to get beautiful results.

``` {r build_tmap, message=FALSE}
## adjust projection to 4326 for both the points an the bounding box objects.
site_sf_4326 <- st_transform(site_sf, 4326)
bb_4326 <- st_transform(bb, 4326)

## plot point on raster using tmap
tm_layout(frame = FALSE) +
tm_shape(bcz) + 
  tm_raster(
    col.legend = tm_legend(title = "Bioclimate"),
    col.scale = tm_scale_categorical(values = 'viridis'),
    col_alpha = 1
    ) + 
tm_shape(st_geometry(site_sf_4326)) + 
tm_dots(fill = "orange", col="orange",  size = 0.1, shape = 20) + 
tm_shape(st_geometry(bb_4326)) + 
  tm_borders(
      lty = 2, 
      col= 'magenta', 
      lwd=2)
```

### Extracting spatial information

Using the simple feature object produced previously from the site data coordinates, you can now extract information from the Biogeographic Region raster layer, using the function `extract()`. These values can then be appended to your site data set using the `merge()` method from the data.table package or your favourite tidy tool.

``` {r extrat_raster, message=FALSE}

site_dt <- data.table(sf::st_drop_geometry(site_sf_4326))
site_sf_4326 <- st_transform(site_sf_4326, crs = crs(bcz))
site_dt$GEnZ_seq <- as.integer(terra::extract(bcz, site_sf_4326, exact = TRUE)$Bioclimate)
site_dt <- merge(site_dt, 
                unique(bcz_class[, .(GEnZ_seq, GEnZname)]), 
                by = "GEnZ_seq",
                all.x=TRUE)
```

### Exploring results

You can have a look at the result using below, with an interactive map to explore what just happened.

``` {r explore_table_map, message = FALSE}
library(DT)
## interactive table with 
DT::datatable(site_dt[ ,.(bms_id, transect_id_serial, GEnZ_seq, GEnZname)], rownames = FALSE, filter = "top", options = list(pageLength = 5, scrollX = T))
## interactive map (crop for efficiency)
crop_bcz <- terra::crop(bcz, st_geometry(bb_4326))
mapview::mapview(crop_bcz) + mapview::mapview(st_transform(site_sf, 4326))
```
