---
title: "Computing site and collated indices"
author: "Reto Schmucki"
date: "29 March 2020"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: "hide"
---

### Generalized Abundance Index (2)

After having successfully computed estimates of the regional flight curve, you can now use this information together with your count data to estimate a total number of butterfly per monitoring site over a season. This will provide you with an estimated abundance for each site that can be used to compute a collated index over the entire regions or a group of sites.

For this task, we will used the flight curve [computed earlier](flight_curve_computing.html) and the `impute_count`, `site_index` and finally `collated_index` functions available in `rbms` package


``` {r package_load}
library(data.table)
library(rbms)
library(ggplot2)
```

Here we will use the count and visit data.

``` {r data_load_and_process}
# s_sp <- "Maniola jurtina"
s_sp <- "Polyommatus icarus"
region_bms <- c("NLBMS", "FRBMS")

b_count_sub <- readRDS("bms_workshop_data/work_count.rds")
m_visit_sub <- readRDS("bms_workshop_data/work_visit.rds")

setnames(m_visit_sub, c("transect_id_serial", "visit_date"), c("SITE_ID", "DATE"))
names(m_visit_sub) <- toupper(names(m_visit_sub))

setnames(b_count_sub, c("transect_id_serial", "visit_date", "species_name"), c("SITE_ID", "DATE", "SPECIES"))
names(b_count_sub) <- toupper(names(b_count_sub))

ts_date <- rbms::ts_dwmy_table(InitYear = 2008,
                               LastYear = 2017,
                               WeekDay1 = "monday")

ts_season <- rbms::ts_monit_season(ts_date, 
                                   StartMonth = 4,
                                   EndMonth = 9, 
                                   StartDay = 1, 
                                   EndDay = NULL,
                                   CompltSeason = TRUE,
                                   Anchor = TRUE,
                                   AnchorLength = 2,
                                   AnchorLag = 2,
                                   TimeUnit = "w")

MY_visit_region <- m_visit_sub[BMS_ID %in% region_bms, ]
ts_season_visit <- rbms::ts_monit_site(MY_visit_region, ts_season)

MY_count_region <- b_count_sub[BMS_ID %in% region_bms, ]
ts_season_count <- rbms::ts_monit_count_site(ts_season_visit, MY_count_region, sp = s_sp)
# ts_flight_curve <- rbms::flight_curve(ts_season_count,
#                                            NbrSample = 500,
#                                            MinVisit = 3,
#                                            MinOccur = 2,
#                                            MinNbrSite = 5,
#                                            MaxTrial = 4,
#                                            GamFamily = "nb",
#                                            SpeedGam = FALSE,
#                                            CompltSeason = TRUE,
#                                            SelectYear = NULL,
#                                            TimeUnit = "w"
#                                            )
ts_flight_curve <- readRDS(file.path("bms_workshop_data", paste(gsub(" ", "_", s_sp), paste(region_bms, collapse="_"), "pheno.rds", sep = "_")))
pheno <- ts_flight_curve$pheno
```

### Impute & Site index

With the phenology and teh observed counts, you can now impute values for missing week counts and adequately compute sites abundance index over the entire monitoring season.

``` {r impute}
impt_counts <- rbms::impute_count(ts_season_count = ts_season_count, 
                                  ts_flight_curve = pheno, 
                                  YearLimit = NULL, 
                                  TimeUnit = "w")

sindex <- rbms::site_index(butterfly_count = impt_counts, MinFC = 0.10)

saveRDS(sindex, file.path("bms_workshop_data", paste( gsub(" ", "_", s_sp), paste(region_bms, collapse="_"), "sindex.rds", sep="_")))
```

### Collated index

Now that we have all each site index, we can collate them together to estimate annual abundance indices for the wider region, using the `collated_index` function in `rbms`. For this example, we will focus on the sites monitored in the Netherlands (NLBMS) and compute a collated index for *Polyommatus icarus* accross the site monitored in that region over the 10-year period.

``` {r collated_index}
bms_id <- "NLBMS"

sindex <- readRDS(file.path("bms_workshop_data", paste( gsub(" ", "_", s_sp), paste(region_bms, collapse="_"), "sindex.rds", sep="_")))

sindex[substr(SITE_ID, 1, 5) == bms_id, ]

co_index <- collated_index(data = sindex[substr(SITE_ID, 1, 5) == bms_id, ], 
                           s_sp = s_sp,
                           sindex_value = "SINDEX",
                           glm_weights = TRUE,
                           rm_zero = TRUE)
co_index$col_index
```

### Bootstrap CI

With this sample of site indices, we can compute the collated indices for multiple combination of sites. Using a bootstrap approach, we can now estimate the confidence interval around the estimated collated index, using the variation observed between monitoring sites. First we randomly draw *n* bootstrap samples, using the function `boot_sample`. This function sample with replacement from a the pool of all sites available. Because sampling is done with replacement, site can occur twice in a specific bootstrap sample and it will be treated as two independent sites (realisations). Here we use a 500 bootstrap, with set.seed() sake of cpu time and repeatability, but you should aim for 1000 sample and more and remove the set.seed() in the random number generator to generate different samples between runs.

``` {r bootstrap_sample, eval = FALSE}
bms_id <- "NLBMS"
set.seed(125361)

bootsample <- rbms::boot_sample(sindex[substr(SITE_ID, 1, 5) == bms_id, ], boot_n = 500)
co_index <- list()

## for progression bar, uncomment the following
pb <- txtProgressBar(min = 0, max = dim(bootsample$boot_ind)[1], initial = 0, char = "*",  style = 3)

for (i in c(0, seq_len(dim(bootsample$boot_ind)[1]))) {
    co_index[[i + 1]] <- rbms::collated_index(data = sindex[substr(SITE_ID, 1, 5) == bms_id, ],
                                              s_sp = s_sp,
                                              sindex_value = "SINDEX",
                                              bootID = i,
                                              boot_ind = bootsample,
                                              glm_weights = TRUE,
                                              rm_zero = TRUE)

    ## for progression bar, uncomment the following
    setTxtProgressBar(pb, i)
}

## collate and append all the result in a data.table format
co_index <- rbindlist(lapply(co_index, FUN = "[[", "col_index"))
co_index$BMS_ID <- bms_id
co_index$SPECIES <- s_sp

co_index_boot <- co_index
saveRDS(co_index_boot, file.path("bms_workshop_data", paste( gsub(" ", "_", s_sp), bms_id, "co_index_boot.rds", sep="_")))
```

### Figure with CI

Now that you have a collated index, and a series bootstrapped collated indices, you can compute a 95% Confidence Interval around your index and construct a figure to illustrate the temporal trend resulting from the annual indices. Here we will rescale and present the indices on a log(index)/log(10) scale. Confidence interval are computed using the 0.025 and 0.975 quantile from the *n* bootstrap sample.  

``` {r collated_index_figure}
bms_id <- "NLBMS"

co_index_boot <- readRDS(file.path("bms_workshop_data", paste( gsub(" ", "_", s_sp), bms_id, "co_index_boot.rds", sep="_")))

boot0 <- co_index_boot[ , mean(COL_INDEX, na.rm=TRUE), by = M_YEAR]
boot0[, LOGDENSITY0 := log(V1) / log(10)]
boot0[, meanlog0 := mean(LOGDENSITY0)]

# Add mean of BOOTi == 0 (real data) to all data
boot_ests <- merge(co_index_boot, boot0[, .(M_YEAR, meanlog0)],
    by = c("M_YEAR")
    )

boot_ests[, LOGDENSITY := log(COL_INDEX) / log(10)]
boot_ests[, TRMOBSCI := LOGDENSITY - meanlog0 + 2]

boot_ests[, TRMOBSCILOW := quantile(TRMOBSCI, 0.025), by = M_YEAR]
boot_ests[, TRMOBSCIUPP := quantile(TRMOBSCI, 0.975), by = M_YEAR]

ylim_ <- c(min(floor(range(boot_ests[TRMOBSCI <= TRMOBSCIUPP & TRMOBSCI >= TRMOBSCILOW, TRMOBSCI]))),
           max(ceiling(range(boot_ests[TRMOBSCI <= TRMOBSCIUPP & TRMOBSCI >= TRMOBSCILOW, TRMOBSCI]))))

plot_col <- ggplot(data = boot_ests[TRMOBSCI <= TRMOBSCIUPP & TRMOBSCI >= TRMOBSCILOW, ], aes(x = M_YEAR, y = TRMOBSCI)) + 
    geom_point(colour = "orange", alpha = 0.2) + ylim(ylim_) +
    geom_line(data = boot_ests[TRMOBSCI <= TRMOBSCIUPP & TRMOBSCI >= TRMOBSCILOW, median(TRMOBSCI, na.rm=TRUE), by = M_YEAR], aes(x = M_YEAR, y = V1), linewidth = 1, colour = "dodgerblue4") +
    geom_hline(yintercept = 2, linetype = "dashed", color = "grey50") +
    scale_x_continuous(minor_breaks = waiver(), limits = c(2005, 2020), breaks = seq(2005, 2020, by = 5)) + 
    xlab("Year") + ylab("Abundance Incices (Log/Log(10))") +
    labs(subtitle = paste(s_sp, bms_id, sep = " "))

plot_col

```