---
title: "Calculating species trends"
author: "Emily Dennis"
date: "31 March 2020"
output: 
    html_document:
     toc: true
     toc_float: true
     code_folding: "hide"
---


Source trend functions required that you can download from [**this link**](workshop_functions.R).

```{r source functions}
source("workshop_functions.R")
```


Read in collated indices for species and BMS of interest
```{r}
# Specify the species - note the underscore
spp <- "Maniola_jurtina"
# Specify the BMS
bms <- "UKBMS"
# Read in the bootstrapped collated indices
co_index <- readRDS(paste0("./bms_workshop_data/", spp, "_co_index_boot.rds"))
# Filter to a single BMS to focus upon
co_index <- co_index[BMS_ID == bms]
co_index
```

Convert to log 10 collated indices (TRMOBS) 
```{r}
co_index[, LOGDENSITY:= log(COL_INDEX)/log(10)][, TRMOBS := LOGDENSITY - mean(LOGDENSITY) + 2, by = .(BOOTi)]
co_index
```

Estimate and classify species trends with a confidence interval based on the bootstraps
```{r}
sp_trend <- estimate_boot_trends(co_index)
sp_trend
```


Plot the species log collated index with bootstrapped 95% confidence interval and linear trend line (in red)
```{r}
# Calculate mean log index for original data
co_index0_mean <- mean(co_index[BOOTi == 0]$LOGDENSITY)
# Derive interval from quantiles
co_index_ci <- merge(co_index[BOOTi == 0, .(M_YEAR, TRMOBS, BMS_ID)],
                        co_index[BOOTi != 0, .(LOWER = quantile(LOGDENSITY - co_index0_mean + 2, 0.025),
                          UPPER = quantile(LOGDENSITY - co_index0_mean + 2, 0.975)), by = .(BMS_ID, M_YEAR)],
                        by=c("BMS_ID","M_YEAR"))


ggplot(co_index_ci, aes(M_YEAR, TRMOBS))+
  theme(text = element_text(size = 14))+
  geom_line()+
  geom_point()+
  geom_ribbon(aes(ymin = LOWER, ymax = UPPER), alpha = .3)+
  geom_smooth(method="lm", se=FALSE, color="red")+
  xlab("Year")+ylab(expression('log '['(10)']*' Collated Index'))+
  ggtitle(paste("Collated index for", gsub("_", " ", spp), "in", bms))

```
